---
title: "01-Syntax, base, functions, loops and data types"
author: "Adi Sarid / adi@sarid-ins.co.il"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goals of this exercise

Familiarize yourself with the basics of R. This exercise encompasses:

   1. Setting variables
   2. Various data types, 
   3. Working with vectors, data.frames
   4. Installing and loading packages
   5. Building functions and base-R iterations



## Setting variables

A variables can be defined using the arrow notation (which we have already seen in the previous lesson `<-`). You can also do the same with `=` but its less common and should not be used (apart from a specific case, in function arguments, which we will discuss later).

```{r setting new variables}
a <- 1
b <- 2
c = 3 # just to show that this works

a
b
c
a*b
b*c

```

Try to run the following code in the console. What is the problem with it, can you fix the code?

```
d <- c(a + b)
```

Do not mix the assignment operator `=` (which I told you not to use), with the test equality operator `==`.

```{r test equality}

a == 1
b == a
b > a

```

Also note the use of logicals:
Please explain what each of the following operators do: `& | ! !=`, you can use the following (and modify it in any way):

```
# TRUE FALSE and the likes
TRUE & FALSE
TRUE | FALSE
TRUE & TRUE
!TRUE
FALSE != TRUE
FALSE == FALSE
```

Try the following code. Bonus points if you can explain what's wrong with it (and why that is).

```{r two is not two}

sqrt2 <- sqrt(2)
sqrt2
2 == sqrt2^2

```

## Data types

R has a number of "basic" data types:

   * Integer
   * Numeric (double)
   * Date (posix)
   * Factors
   * Logicals

You can use `c()`, `rbind()`, `cbind()` to piece values together into vectors or more complex structures.
Run the following code.

```{r}
integer_example <- 10L
integer_example
numeric_example <- pi # pi is a reserved word...
numeric_example
character_examples <- "hello world"
character_examples
date_example <- as.Date("2018-10-01")
date_example
factor_example <- as.factor(c("big", "big", "small", "medium", "small", "big", "bigger"))
factor_example
summary(factor_example)
logical_example <- c(TRUE, TRUE, FALSE, TRUE)
logical_example
```


***

Using the `c()` command try to piece together the `logical_example` with the `factor_example`, i.e. (replace the `???` with something else):

```
c(logical_example, ???)
```
What happend to the factor vector? does the resulting vector make sense?

Do the same with the `date_example` and the `factor_example`. What happend now? What precautions would you take when working with factors?

***

## data frames

Data frames are a more complex structure which contains mixed data. R comes bundled with a number of "classical" data frames. Try the following:

```
mtcars
iris
?mtcars
?iris
```

What types are the variables (columns) in each of these data sets? (double/factor/date/logical/integer/character)

## Packages

An R package is a bundle of functions which share a common goal or vision. So far, we've been using base-r. The `tidyverse` packages is a package of packages. We will be working a lot with it. Let's try to load `tidyverse`.

```
library(tidyverse)
```

Did that work? if you got an error message you might need to install it. The following code will download and install the tidyverse. Be warned, this takes long.

```
install.packages("tidyverse")
```

Now if you installed the package, try to load it again `library(tidyverse)`. To use a function after you loaded a packages you can call `function_name(arg1 = ..., arg2 = ..., ...)`. Use `glimpse` to verify your answers for the previous questions (what types are the variables in mtcars and iris):

```{r tidyverse}
library(tidyverse)

glimpse(iris)
glimpse(mtcars)

```

Use the function `count` to answer:

How many flower-types are there in `iris`?
How many cylinder values are there in `mtcars`?

```
count(iris, Species)
count(???, cyl)

```

Later on, we will learn some more convinient ways to answer such questions.


## Functions and iterations - intermediate exercise

We will discuss some base-R iterations, however, **in real situations you should do all in your power to avoid base-r loops!**.

In the following exercise you will build a function which computes the Fibonacci series (0, 1, 1, 2, 3, 5, 8, 13, 21,...), and a loop which does the same. You will compare their runtime using `bench::mark()`.

WARNING:
This might feel like a **relatively complex** exercise if you're not fluent in programming, and it's not directly related to data analysis. The reasone I am giving you this exercise is that it is a great exercise to reherse elements we were discussing, in one single exercise.
   
   1. Functions and the concept of recursion (a function calling itself)
   2. Base-r loops
   3. Conditionals (`if...else if...else`).

First, if you don't know what the Fibonacci series is, go to [wikipedia, Fibonacci number](https://en.wikipedia.org/wiki/Fibonacci_number) and read about it (just the intro, should suffice).

Complete the following function so that a call to the function will generate the n^th^ Fibonnacci number. Replace the `???`.
Rows which start with the hash sign `#` are comments and will be ignored.

Also, if you never heard the term "recursion" up until today, you might want to start with the second function `fib_loop`, and then think about the first one `fibonnaci`.

``` {r} 
fibonnacci <- function(n){
  if (n == 0) {
    # starting condition for F_0 
    return(0)
  } else if (n == 1) {
    # starting condition for F_1
    return(1)
  } else {
    # use recursion to calculate the number
    return(fibonnacci(n-1)+fibonnacci(n-2))
  }
}

fibonnacci(30)

```

An alternative way to compute the Fibonnacci is via loops. Complete the following code:

``` {r}
fib_loop <- function(n){
len <- n
fibvals <- numeric(len)
fibvals[1] <- 1
fibvals[2] <- 1
for (i in 3:len) { 
   fibvals[i] <- fibvals[i-1]+fibvals[i-2]
} 
return( tail(fibvals,1))
}
fib_loop(30)

```

Check that your are getting consistent results. Now, compare the two functions using:

```
install.packages(bench) # if it is not installed
bench::mark(fibonnacci(30), fib_loop(30))
```

Which method is quicker?

Note the use of `::`, I didn't mention this earlier, but instead of loading the package entirely `library(bench)` we're just calling the function `mark` from packages`bench` directly, using the double `::`.