---
title: "Test Exercises - Examples"
author: "Adi"
date: "Feb 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

The test will be comprised of five (5) questions, which will cover the following topics:

   * Visualizations (1 question)
   
   * Understanding model output (1 question)
   
   * Hypothesis testing (2-3 questions): 
   
      * Means
      
      * Distributions
      
      * Regression
      
      * ANOVA

   * Theoretical (up to 1 question)
   


## Exercise 1 - Multiple Linear Regression

The `state.x77` dataset contains various statistics on US states (the 70s). Specifically, it contains the `Income` per capita, `Frost` mean number of days with minimum temprature below freezing, `Illiteracy`, `Murder` murder rate per 100,000 population, `HS Grad` high-school graduates (%).

Two linear regression models were fit to predict the per capita income, the second fit contains `HS Grad` which is not included in the first fit.

```{r lm fits}
library(jtools)
states <- as.data.frame(state.x77)
fit <- lm(Income ~ Frost + Illiteracy + Murder, data = states)
fit2 <- lm(Income ~ Frost + Illiteracy + Murder + `HS Grad`, data = states)
export_summs(fit, fit2, scale = FALSE)
```

**Question 1:** Ignoring the intercept, what are the significant coefficients in each of the models?

**Question 2:** How does an increase of 1 unit in `HS Grad` influence the `Income` variable?

**Question 3:** The following density charts show the distribution of the coefficients estimates with a 95% confidence interval. What does it mean when a density chart does not include the dashed line?

```{r coefficients plot}
plot_summs(fit, fit2, scale = TRUE, plot.distributions = TRUE)
```

**Question 4:** Based on the results, what would you say about the relationship between `Illiteracy` and `HS Grad`, i.e., do they correlate? and if so, would you expect a positive or negative correlation? why?

## Exercise 2 - Axis and Outliers

In `ggplot2` there are a few functions that can be used to control the axis. Two such examples are `coord_cartesian` and `lims`. Look at the following three charts. They are boxplots plotting the data of drivers killed between 1969-1984. 

**Question 1:** Explain the difference in the results between `coord_cartesian` and `lims`. What function would you use for:

   1. Zooming-into a chart.
   2. Omitting observations and analyzing a chart after the omission.

```{r coord cartesian versus lims, fig.dim=c(4,4)}

ggplot(as_tibble(Seatbelts), aes(y = DriversKilled, x = "")) + 
  geom_boxplot()

ggplot(as_tibble(Seatbelts), aes(y = DriversKilled, x = "")) + 
  geom_boxplot() + 
  coord_cartesian(ylim = c(110, 130))

ggplot(as_tibble(Seatbelts), aes(y = DriversKilled, x = "")) + 
  geom_boxplot() + 
  lims(y = c(110, 130))

```

**Question 2:** Would observation A be classified as an outlier when examined using two boxplots (one for the x-axis and one for the y-axis)?

```{r would this be classified as an outlier?}

```