---
title: "Where no R has gone before - Analyzing Star Trek scripts"
author: "Adi Sarid"
date: "12 1 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

In this markdown, I analyze [this](https://www.kaggle.com/gjbroughton/start-trek-scripts/) data of startrek scripts.

## Goals

My main goal is to demonstrate various elements which I expect to see in the project, while allowing the students to ask questions as we build the analysis in real time together.

Here are some related research questions which we can examine:

   * What is the number of words/script lines spoken by each starfleet captain?
   * What is the difference between a captain, number.1, and other characters?
   * What is the relationship between sentence length (number of words) and starfleet captains?
   * Is there a gender bias in different series/episodes?
   * What the number of words per episode? Is it randomely distributed?

# Data Import and Tidying

The source is in json files - non tabular data which is hard to work with. Hence, we must first turn it into a tidyformat.  data was read from the json files and put into this tidy format.

```{r read startrek script data}
suppressWarnings(suppressMessages(library(tidyverse)))
trek <- read_csv("https://github.com/adisarid/startrek_plumber_api/blob/master/raw_data/characters_words.csv?raw=true")
glimpse(trek)
```

# Transformation, Visualization, and Modelling

## Which starfleet captains appear in the data?

First, let's determine who are the starfleet captains which appear in the data.

```{r starfleet captain}
trek %>% 
  group_by(character, series) %>% 
  summarize(num_words = sum(total_words)) %>% 
  arrange(desc(num_words))
```

The starfleet captains which appear are PICARD, JANEWAY, KIRK, SISCO, and ARCHER. Now we can filter the data and examine them.

# Conclusions