---
title: "Lab - food consumption and carbon footprint"
author: "Adi Sarid"
date: "`r Sys.Date()`"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(learnr)
food_consumption <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-18/food_consumption.csv')
```

In this lab we cover the data science work flow. The lab will walk you through the steps performed in a data science project:

   * Data import
   * Data tidying
   * Transformation <-> Visualization <-> Modelling
   
The lab will also cover the theoretic elements we covered such confidence intervals and hypothesis tests. This lab is to be performed in groups of 3 (i.e., zoom break rooms).

# First exercise - open up a new project.

First, open up a new project. To do this, in RStudio go to:

   * File -> New Project -> New Directory -> New Project.

Provide your project directory name (under directory name), and click ok. Note that everything will close and RStudio will open up in a clean window. But don't worry, you can always view this file by visiting [this link](https://sarid.shinyapps.io/food_consumption_lab/).

Once a new RStudio instance has opened with a clean window of your new project, open up a new RMarkdown, which you will use to answer your questions to this lab by:

   * File -> New file -> R Markdown...

Give your RMarkdown file a name and use the html outupt type. In your new RMarkdown file you can delete lines 1-10 and delete everything else in the file (lines 11-31). Try to knit it by clicking ctrl+k. 

Now we are ready to do some analysis.

# Second exercise - get to know your data

In this lab we are going to analyze food consumption data from *tidytuesday*. You can read about tidytuesday [here](https://github.com/rfordatascience/tidytuesday) - it's a github repository which is updated every Tuesday with data freely available for analyzing and sharpening your data analysis skills. 
Today we will use this dataset: [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-02-18/readme.md).

In groups, find the original source of the data (the nu3.de website within the links) and discuss: 

   * What is the origin of the data?
   
   * Would you consider the data reliable/trust worthy?
   
   * How was the carbon footprint computed for each food type and country?

Using the `read_csv` function from the `readr` package, read the food consumption data. 

   * Use the following functions to understand how the data is arranged, for example with `glimpse`, `head` or just typing the  `food_consumption` name. In an interactive rstudio session you can also use `View`. 
   
   * Comparing the file you read with the original table, in what sense the file you read is more "tidy"?

```{r read_the_data, exercise=TRUE}
food_consumption <- readr::read_csv(...)
```

```{r read_the_data-hint-1}
food_consumption <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-18/food_consumption.csv')
food_consumption
```

```{r read_the_data-hint-2}
glimpse(...)
```

```{r read_the_data-hint-3}
head(...)
```

```{r read_the_data-solution}
food_consumption <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-18/food_consumption.csv')
food_consumption
glimpse(food_consumption)
head(food_consumption)

```

***

**Checkpoint** - make sure you summarize your answers in a convinient way, we are going to discuss them in class.

***

# Third exercise - visualization, and descriptive statistics of food consumption

In the group, discuss: 

   * What influences the consumption of various goods (e.g., pork, beef, fish, wheat, etc.). In which countries would you expect to find a low or high consumption of specific products?
   
   * Write down an equivalent formula for the variance. This is a theoretic part, you can write your answer in RMarkdown by enclosing it with the $ characters like this. RMarkdown interprets such test as LaTeX and creates a formula:

$$\sigma^2=\operatorname{Var}(X)=E\left[(X-E[X])^2\right]=\ldots?$$

   * Write down two estimates for the variance $\hat{\sigma}^2$ and $s^2$. Which would you prefer to use and why?
   
   * What plot(s) would you use to visualize the distribution of consumption of each food type? you can use the `ggplot2` cheatsheet to consider this (there is more than one option). Once you reach a conclusion - create the chart(s).

```{r food_consumption_distribution, exercise=TRUE}

```

```{r food_consumption_distribution-hint-1}
ggplot(food_consumption, ...) + 
   ...
```

```{r food_consumption_distribution-hint-2}
ggplot(food_consumption, aes(x = ..., y = ...)) + 
   geom_...()
```
   
```{r food_consumption_distribution-hint-3}
ggplot(food_consumption, aes(x = food_category, y = ...)) + 
   geom_boxplot() +
   coord_flip()
```
   
```{r food_consumption_distribution-solution}
# Option 1:
ggplot(food_consumption, aes(x = food_category, y = consumption)) + 
   geom_boxplot() + coord_flip()

# Option 2:
ggplot(food_consumption, aes(x = consumption)) + 
   geom_histogram() + 
   facet_wrap(~food_category)

# Option 3:
ggplot(food_consumption, aes(x = consumption)) + 
   geom_density() + 
   facet_wrap(~food_category, scales = "free")
```



   * Based on the plot, what products have the highest (or lowest) variance? what is the meaning of having a high (or low) variance in this context of food consumption?
   
   * Verify this by computing the consumption standard deviation of each product, also add to your computation the mean, median and 1st and 3rd quartiles. You can use `group_by` and `summarize` for this.

```{r summary_stats, exercise=TRUE}

food_descriptives <- food_consumption %>% 
   ...

```

```{r summary_stats-hint-1}
food_descriptives <- food_consumption %>% 
   group_by(...) %>% 
   ...
```  

```{r summary_stats-hint-2}
food_descriptives <- food_consumption %>% 
   group_by(food_category) %>% 
   summarize(...)
```
 
```{r summary_stats-hint-3}
food_descriptives <- food_consumption %>% 
   group_by(food_category) %>% 
   summarize(mean = ...,
             q1 = ...,
             q2 = ...,
             q3 = ...,
             sd = ...)
```
 
```{r summary_stats-hint-4}
food_descriptives <- food_consumption %>% 
   group_by(food_category) %>% 
   summarize(mean = mean(consumption),
             q1 = quantile(..., probs = 0.25),
             q2 = median(...),
             q3 = quantile(..., probs = ...),
             sd = sd(...))
```
 
```{r summary_stats-solution}
food_descriptives <- food_consumption %>% 
   group_by(food_category) %>% 
   summarize(mean = mean(consumption),
             q1 = quantile(consumption, probs = 0.25),
             q2 = median(consumption),
             q3 = quantile(consumption, probs = 0.75),
             sd = sd(consumption)) %>% 
   arrange(desc(mean))
food_descriptives
```

```{r create food_descriptives for later, include=FALSE}
food_descriptives <- food_consumption %>% 
   group_by(food_category) %>% 
   summarize(mean = mean(consumption),
             q1 = quantile(consumption, probs = 0.25),
             q2 = median(consumption),
             q3 = quantile(consumption, probs = 0.75),
             sd = sd(consumption)) %>% 
   arrange(desc(mean))
```

***

**Checkpoint** - together in class, we're going to discuss and solve the exercise so far.

***

# Fourth exercise - modelling - confidence intervals and hypothesis tests of consumption

In this part we will create a number of confidence intervals. Follow these steps in order to answer this question:

   * First, decide what kind of confidence interval is to be used (what statistic are you using) to compare the consumption of food categories?
   
```{r test-statistic, echo=FALSE}
question("What test statistic should we use to create a confidence interval around the mean consumption of food categories?",
         answer("$Z$"),
         answer("$T$", correct = TRUE),
         answer("$\\chi^2$"),
         answer("Either $Z$ or $T$, it doesn't matter which"),
         allow_retry = T)
```
   
   * Write it down as a formula:

$$?_{\text{df}=n-1} = \frac{??? - ???}{???/\sqrt{???}}$$

   * And the confidence interval is therefore:

$$\\mu\in?\pm t_{\alpha/2,n-1}?/\sqrt{?}$$

   * Use the tibble you created in the last step of the previous part (`food_descriptives`), to create a confidence interval for all the food categories, with $\alpha=0.05$. Use the following code.

```{r confidence_intervals_t_statistic, exercise=TRUE}

# First find the relevant t for the chosen alpha and the relevant degrees of freedom
t0.05_129 <- ... # you should consider one of: dt, pt, qt, rt

# Now use it on the tibble we computed
food_descriptives %>% 
   ...

```

```{r confidence_intervals_t_statistic-hint-1}

# First find the relevant t for the chosen alpha and the relevant degrees of freedom
t0.05_129 <- qt(p = ..., df = ...) # you should consider one of: dt, pt, qt, rt

# Now use it on the tibble we computed
food_descriptives %>% 
   ...

```

```{r confidence_intervals_t_statistic-hint-2}

# First find the relevant t for the chosen alpha and the relevant degrees of freedom
t0.05_129 <- qt(p = 0.05/2, df = 129)

# Now use it on the tibble we computed
food_descriptives %>% 
   ...

```

```{r confidence_intervals_t_statistic-hint-3}

# First find the relevant t for the chosen alpha and the relevant degrees of freedom
t0.05_129 <- qt(p = 0.05/2, df = 129)

# Now use it on the tibble we computed
food_descriptives %>% 
   mutate(ci_lower_bound = ...,
          ci_upper_bound = ...) %>% 
   select(food_category, ci_lower_bound, mean, ci_upper_bound)

```

```{r confidence_intervals_t_statistic-hint-4}

# First find the relevant t for the chosen alpha and the relevant degrees of freedom
t0.05_129 <- qt(p = 0.05/2, df = 129)

# Now use it on the tibble we computed
food_descriptives %>% 
   mutate(ci_lower_bound = mean + t0.05_129*sd/sqrt(130),
          ci_upper_bound = ...) %>% 
   select(food_category, ci_lower_bound, mean, ci_upper_bound)

```

```{r confidence_intervals_t_statistic-solution}

# First find the relevant t for the chosen alpha and the relevant degrees of freedom
t0.05_129 <- qt(p = 0.05/2, df = 129)

# Now use it on the tibble we computed
food_descriptives %>% 
   mutate(ci_lower_bound = mean + t0.05_129*sd/sqrt(130),
          ci_upper_bound = mean - t0.05_129*sd/sqrt(130)) %>% 
   select(food_category, ci_lower_bound, mean, ci_upper_bound)

```

   * Check your result using the `t.test` function for Fish.

```{r fish_ci, exercise=TRUE}

fish_vector <- food_consumption %>% 
   filter(...) %>% 
   pull(...)

t.test(...)

```

```{r fish_ci-hint-1}

fish_vector <- food_consumption %>% 
   filter(food_category == "...") %>% 
   pull(consumption)

t.test(...)

```

```{r fish_ci-hint-2}

fish_vector <- food_consumption %>% 
   filter(food_category == "Fish") %>% 
   pull(consumption)

t.test(...)

```

```{r fish_ci-solution}

fish_vector <- food_consumption %>% 
   filter(food_category == "Fish") %>% 
   pull(consumption)

t.test(fish_vector)

```

```{r fish ci, include=FALSE}

fish_vector <- food_consumption %>% 
   filter(food_category == "Fish") %>% 
   pull(consumption)

t.test(fish_vector)

```

   * Formulate a hypothesis test which examines the expected consumption of Fish vs. Pork (with $H_0$ and $H_1$). Check all that can be used:
   
```{r hypothesis-test-formulation-question, echo=FALSE}
question("What hypothesis tests can be used to compare pork and fish consumption?",
         allow_retry = TRUE,
         random_answer_order = TRUE,
         type = "checkbox",
         answer("$H_0: \\mu_{\\text{pork}} = \\mu_{\\text{fish}}, H_1:\\mu_{\\text{pork}} \\neq \\mu_{\\text{fish}}$", correct = TRUE),
         answer("$H_0: \\mu_{\\text{pork}} = \\mu_{\\text{fish}}, H_1:\\mu_{\\text{pork}} > \\mu_{\\text{fish}}$", correct = TRUE),
         answer("$H_0: \\mu_{\\text{pork}} \\neq \\mu_{\\text{fish}}, H_1:\\mu_{\\text{pork}} = \\mu_{\\text{fish}}$", correct = FALSE),
         answer("$H_0: \\mu_{\\text{pork}} < \\mu_{\\text{fish}}, H_1:\\mu_{\\text{pork}} > \\mu_{\\text{fish}}$", correct = FALSE),
         answer("$H_0: \\mu_{\\text{pork}} \\neq \\mu_{\\text{fish}}, H_1:\\mu_{\\text{pork}} < \\mu_{\\text{fish}}$", correct = FALSE))
```
   
   * Assuming that the variance of pork and fish consumption is the same, what test statistic would you use for this hypothesis?
   
$$T=\\frac{? - ? - (\\mu_1 - \\mu_2)}{S_p\\sqrt{? + ?}}$$

$$S_p = ?$$

With df$=?$.

```{r hypothesis_test_formulation_paired_unpaired, echo=FALSE}
question("Is this test paired or unpaired?",
         allow_retry = TRUE,
         random_answer_order = TRUE,
         type = "radio",
         answer("Paired", correct = FALSE),
         answer("Unpaired", correct = TRUE),
         answer("Irrelevant", correct = FALSE),
         answer("Can be paired or unpaired - it doesn't matter", correct = FALSE))
```

   * Conduct the test by computing the test statistic and its p-value. You can do this either directly or with the `t.test` function. However you prefer. If you are using the `t.test`, note that you have to set the `var.equal` argument (to what?).

```{r compare_fish_and_pork, exercise=TRUE}
pork_vector <- ...

t.test(x = ..., y = ..., var.equal = ...)

```

```{r compare_fish_and_pork-hint-1}
pork_vector <- food_consumption %>% 
   filter(food_category == "Pork") %>% 
   pull(consumption)

t.test(x = ..., y = ..., var.equal = ...)
```

```{r compare_fish_and_pork-solution}
pork_vector <- food_consumption %>% 
   filter(food_category == "Pork") %>% 
   pull(consumption)

t.test(x = fish_vector, y = pork_vector, var.equal = T)

```
   
   * How would you have conducted the test if the variances were assumed to be unequal?

***

**Checkpoint** - solving this exercise together in class.

***

# Fifth exercise - visualizing the relationship between meet products and vegan products

In this final part, we're going to use visualizations to examine the relationship between meat/dairy and vegan products.

   * Reclassify all the `food_category` into two types of products: meet/dairy versus vegan. You can use the definition of the following tibble, along with the `left_join` function, but you will probably need to read about it in the documentation. Another option is to use another function called `case_when` or `recode_factor`.
   
```{r food_reclassification, exercise=TRUE}
food_types <- tribble(~food_category, ~food_type,
                      "Beef", "Meat/Dairy",
                      "Eggs", "Meat/Dairy",
                      "Fish", "Meat/Dairy",
                      "Lamb & Goat", "Meat/Dairy",
                      "Milk - inc. cheese", "Meat/Dairy",
                      "Pork", "Meat/Dairy",
                      "Poultry", "Meat/Dairy",
                      "Rice", "Vegan",
                      "Soybeans", "Vegan",
                      "Wheat and Wheat Products", "Vegan",
                      "Nuts inc. Peanut Butter", "Vegan")

food_consumption_reclassified <- food_consumption %>% 
   ...

```

```{r food_reclassification-hint-1}
food_types <- tribble(~food_category, ~food_type,
                      "Beef", "Meat/Dairy",
                      "Eggs", "Meat/Dairy",
                      "Fish", "Meat/Dairy",
                      "Lamb & Goat", "Meat/Dairy",
                      "Milk - inc. cheese", "Meat/Dairy",
                      "Pork", "Meat/Dairy",
                      "Poultry", "Meat/Dairy",
                      "Rice", "Vegan",
                      "Soybeans", "Vegan",
                      "Wheat and Wheat Products", "Vegan",
                      "Nuts inc. Peanut Butter", "Vegan")

food_consumption_reclassified <- food_consumption %>% 
   left_join(...)

```

```{r food_reclassification-solution}
food_types <- tribble(~food_category, ~food_type,
                      "Beef", "Meat/Dairy",
                      "Eggs", "Meat/Dairy",
                      "Fish", "Meat/Dairy",
                      "Lamb & Goat", "Meat/Dairy",
                      "Milk - inc. cheese", "Meat/Dairy",
                      "Pork", "Meat/Dairy",
                      "Poultry", "Meat/Dairy",
                      "Rice", "Vegan",
                      "Soybeans", "Vegan",
                      "Wheat and Wheat Products", "Vegan",
                      "Nuts inc. Peanut Butter", "Vegan")

food_consumption_reclassified <- food_consumption %>% 
   left_join(food_types)

```

```{r food_reclassification actual run, include=FALSE}
food_types <- tribble(~food_category, ~food_type,
                      "Beef", "Meat/Dairy",
                      "Eggs", "Meat/Dairy",
                      "Fish", "Meat/Dairy",
                      "Lamb & Goat", "Meat/Dairy",
                      "Milk - inc. cheese", "Meat/Dairy",
                      "Pork", "Meat/Dairy",
                      "Poultry", "Meat/Dairy",
                      "Rice", "Vegan",
                      "Soybeans", "Vegan",
                      "Wheat and Wheat Products", "Vegan",
                      "Nuts inc. Peanut Butter", "Vegan")

food_consumption_reclassified <- food_consumption %>% 
   left_join(food_types)

```

   * In the result you got, summarize the data such that each country will appear only twice (once for Meat/Dairy and once for Vegan consumption values), with the overall consumption for that type. Select only the `consumption` and `food_type` values. You should be using the functions `group_by` and `summary`.
   
```{r reclassification_summary, exercise=TRUE}
food_consumption_summarised <- food_consumption_reclassified %>% 
   ...
```

```{r reclassification_summary-hint-1}
food_consumption_summarised <- food_consumption_reclassified %>% 
   group_by(..., ...) %>% 
   summarize(consumption = sum(...))
```

```{r reclassification_summary-hint-2}
food_consumption_summarised <- food_consumption_reclassified %>% 
   group_by(country, ...) %>% 
   summarize(consumption = sum(...))
```

```{r reclassification_summary-hint-3}
food_consumption_summarised <- food_consumption_reclassified %>% 
   group_by(country, food_type) %>% 
   summarize(consumption = sum(...))
```

```{r reclassification_summary-solution}
food_consumption_summarised <- food_consumption_reclassified %>% 
   group_by(country, food_type) %>% 
   summarize(consumption = sum(consumption))
```

```{r create food consumption summarized, include=FALSE}
food_consumption_summarised <- food_consumption_reclassified %>% 
   group_by(country, food_type) %>% 
   summarize(consumption = sum(consumption))
```

   * We would like to create a chart in which each country is a point, the x axis is Meat/dairy and the y-axis is Vegan consumption. What kind of transformations would you need to do on the previous tibble to prepare it for such a plot?
   
   * Try to use `pivot_wider` in order to make that transformation, and use `ggplot` to create the chart.
   
```{r consumption_chart, exercise=TRUE}
food_consumption_summarised %>% 
   ungroup() %>% 
   select(country, consumption, food_type) %>% 
   pivot_wider(...) %>% 
   ggplot(...) + 
   ...
```

```{r consumption_chart-hint-1}
food_consumption_summarised %>% 
   ungroup() %>% 
   select(country, consumption, food_type) %>% 
   pivot_wider(names_from = food_type, values_from = ...) %>% 
   ggplot(...) + 
   ...
```

```{r consumption_chart-hint-2}
food_consumption_summarised %>% 
   ungroup() %>% 
   select(country, consumption, food_type) %>% 
   pivot_wider(names_from = food_type, values_from = consumption) %>% 
   ggplot(aes(x = `Meat/Dairy`, y = ...)) + 
   geom_...()
```

```{r consumption_chart-hint-3}
food_consumption_summarised %>% 
   ungroup() %>% 
   select(country, consumption, food_type) %>% 
   pivot_wider(names_from = food_type, values_from = consumption) %>% 
   ggplot(aes(x = `Meat/Dairy`, y = Vegan)) + 
   geom_...()
```

```{r consumption_chart-solution}
food_consumption_summarised %>% 
   ungroup() %>% 
   select(country, consumption, food_type) %>% 
   pivot_wider(names_from = food_type, values_from = consumption) %>% 
   ggplot(aes(x = `Meat/Dairy`, y = Vegan)) + 
   geom_point()
```

   * Can you identify any relationships between the two variables? (consumption of Meat/Dairy versus consumption of Vegan food?)


***

**Checkpoint** - Solve exercise together in class.

***