---
title: "Linear regression"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
```

## Background

In this tutorial we're going to use visualization and linear regression in order to **explore the relationship between tree diameter and age**.

This tutorial was developed by Adi Sarid, as part of an introduction to statistics and data analysis with R. The tutorial is provided under the [cc-by-nc-sa license](https://creativecommons.org/licenses/by-nc-sa/4.0/). Source code is available in [this repo](https://github.com/adisarid/intro_statistics_R) under the labs folder.

We'll be using the [San Francisco Trees](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-28/readme.md) data set from tidytuesday.

## Familiarize yourself with the data

First, let's read the data and get familiar with it. The original dataset contains many records of trees. Follow [this link](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-28/readme.md) to read more. For the purposes of this tutorial we have taken a smaller portion of it and put it in `sf_raw`. It contains many trees spread over the streets of San Francisco. `glimpse` on it, you will notice that there is a variable called `date`, discuss in the group:

   * What is the meaning of `date`? 
   * How can we compute the tree's `age` using the date?
   * Implement this and put it in a new tibble called `sf_trees`

```{r prep data which s going to be used, include=FALSE}
sf_raw <- read_csv('https://raw.githubusercontent.com/adisarid/intro_statistics_R/master/labs/data/sf_trees.csv') 

sf_trees <- sf_raw %>% 
  mutate(age = 2020 - lubridate::year(date))
```

```{r initial_read, exercise=TRUE}

glimpse(sf_raw)

sf_trees <- sf_raw %>%  
  mutate(age = ...)

```

```{r initial_read-hint-1}

sf_trees <- sf_raw %>% 
  mutate(age = 2020 - lubridate::...)

```

```{r initial_read-solution}

sf_trees <- sf_raw %>% 
  mutate(age = 2020 - lubridate::year(date))

```

Create a tibble with the frequency of each tree species and sorted in descending order of number of appearances, then create a plot showing the frequency of the 20 most common tree species (also sorted by number of appearances).

```{r plot_tree_type_frequency, exercise=TRUE}
# You can do either one of the following (or both)

# The tibble:
sf_type_frequency <- ... %>% 
  ...(...) %>% 
  arrange(desc(...)) %>% 
  mutate(species = fct_inorder(...))

sf_type_frequency

# The plot:
ggplot(sf_type_frequency %>% 
         slice(...), 
       aes(y = ..., x = ...)) + 
  geom_...

```

```{r plot_tree_type_frequency-hint-1}
# You can do either one of the following (or both)

# The tibble:
sf_type_frequency <- sf_trees %>% 
  count(species) %>% 
  arrange(desc(...)) %>% 
  mutate(species = fct_inorder(...))

sf_type_frequency

# The plot:
ggplot(sf_type_frequency %>% 
         slice(...), 
       aes(y = ..., x = ...)) + 
  geom_...

```

```{r plot_tree_type_frequency-hint-2}
# You can do either one of the following (or both)

# The tibble:
sf_type_frequency <- sf_trees %>% 
  count(species) %>% 
  arrange(desc(n)) %>% 
  mutate(species = fct_inorder(...))

sf_type_frequency

# The plot:
ggplot(sf_type_frequency %>% 
         slice(...), 
       aes(y = ..., x = ...)) + 
  geom_...

```

```{r plot_tree_type_frequency-hint-3}
# You can do either one of the following (or both)

# The tibble:
sf_type_frequency <- sf_trees %>% 
  count(species) %>% 
  arrange(desc(n)) %>% 
  mutate(species = fct_inorder(species))

sf_type_frequency

# The plot:
ggplot(sf_type_frequency %>% 
         slice(1:...), 
       aes(y = ..., x = ...)) + 
  geom_...

```

```{r plot_tree_type_frequency-hint-4}
# You can do either one of the following (or both)

# The tibble:
sf_type_frequency <- sf_trees %>% 
  count(species) %>% 
  arrange(desc(n)) %>% 
  mutate(species = fct_inorder(species))

sf_type_frequency

# The plot:
ggplot(sf_type_frequency %>% 
         slice(1:20), 
       aes(y = species, x = ...)) + 
  geom_...

```

```{r plot_tree_type_frequency-solution}
# You can do either one of the following (or both)

# The tibble:
sf_type_frequency <- sf_trees %>% 
  count(species) %>% 
  arrange(desc(n)) %>% 
  mutate(species = fct_inorder(species))

sf_type_frequency

# The plot:
ggplot(sf_type_frequency %>% 
         slice(1:20), 
       aes(y = species, x = n)) + 
  geom_col()

```

```{r invalid-category}
quiz(
  question_text(
    "Do you see any category which is not actually a tree species? write down its name.",
    allow_retry = T,
    answer("Tree(s) ::", correct = T),
    answer("Tree(s)::", correct = T),
    answer("Tree(s) :: ", correct = T),
    answer("Tree(s):: ", correct = T),
    answer("Tree(s)", correct = T)
  )
)
```

## Linear regression

Discuss in the group: 

   1. What variables in the data set, other than age, might influence or be related to the tree's diameter at breast height (`dbh`)?

   2. What are the difficulties involved in using them in a linear regression model?

```{r dependent-and-independent-variables}
quiz(
  question("In a model which examines the relationship between the tree's age and diameter, what would you define as the dependent variable $y$ and what would you define as the independent variable $x$?", allow_retry = T, random_answer_order = T,
    answer("$y$ is the diameter, $x$ is the age", correct = TRUE),
    answer("$y$ is the age, $x$ is the diameter"),
    answer("It doesn't really matter because mathematically it's the same"),
    answer("You can't use the two variables in a linear regression model, because both are continuous. First, one of them must be grouped"),
    answer("Depending on what other variables are also used in the model")
  )
)
```

The following plot which depicts the relationship between age and diameter of the two species: *Brisbane Box* and *Swamp Myrtle*. Under the plot there is a code chunk which was used to create the plot. Complete the blanks in the code chunk to recreate the plot.

```{r brisbane_box-and-swamp_myrtle_plot_answer, echo = FALSE}
sf_trees %>% 
  filter(str_detect(species, "Brisbane Box|Swamp Myrtle")) %>% 
  ggplot(aes(x = age, y = dbh, color = species)) + 
  geom_point() + 
  stat_smooth(method = "lm", formula = y ~ x) + 
  theme(legend.position = "top")
```

```{r brisbane_box-and-swamp_myrtle_plot, exercise=TRUE}
sf_trees %>% 
  filter(...) %>% 
  ggplot(...) + 
  geom_... + 
  stat_... + 
  theme(legend.position = "top")
```

```{r brisbane_box-and-swamp_myrtle_plot-hint-1}
sf_trees %>% 
  filter(str_detect(species, "Brisbane Box|Swamp Myrtle")) %>% 
  ggplot(...) + 
  geom_... + 
  stat_... + 
  theme(legend.position = "top")
```

```{r brisbane_box-and-swamp_myrtle_plot-hint-2}
sf_trees %>% 
  filter(str_detect(species, "Brisbane Box|Swamp Myrtle")) %>% 
  ggplot(aes(x = age, y = dbh, color = species)) + 
  geom_... + 
  stat_... + 
  theme(legend.position = "top")
```

```{r brisbane_box-and-swamp_myrtle_plot-hint-3}
sf_trees %>% 
  filter(str_detect(species, "Brisbane Box|Swamp Myrtle")) %>% 
  ggplot(aes(x = age, y = dbh, color = species)) + 
  geom_point() + 
  stat_smooth(method = ...) + 
  theme(legend.position = "top")
```

```{r brisbane_box-and-swamp_myrtle_plot-solution}
sf_trees %>% 
  filter(str_detect(species, "Brisbane Box|Swamp Myrtle")) %>% 
  ggplot(aes(x = age, y = dbh, color = species)) + 
  geom_point() + 
  stat_smooth(method = "lm", formula = y ~ x) + 
  theme(legend.position = "top")
```

Discuss in the group:

   1. What happens when you place the `color` aesthetics in the `ggplot2` function versus the `geom_...()``` function?
   2. What is the difference between the two species, in terms of how age affects diameter?
   
Using the `lm` function we have shown in class build two linear regression models, for each species and show a summary of the two models.

```{r tree_model_regression, exercise = TRUE}
brisbane <- sf_trees %>% 
  ...

myrtle <- sf_trees %>% 
  ...

brisbane_lm <- lm(...)
myrtle_lm <- lm(...)

summary(brisbane_lm)
summary(...)
```

```{r tree_model_regression-hint-1}
brisbane <- sf_trees %>% 
  filter(...)

myrtle <- sf_trees %>% 
  filter(...)


brisbane_lm <- lm(...)
myrtle_lm <- lm(...)

summary(brisbane_lm)
summary(...)
```

```{r tree_model_regression-hint-2}
brisbane <- sf_trees %>% 
  filter(str_detect(species, "Brisbane Box"))

myrtle <- sf_trees %>% 
  filter(...)


brisbane_lm <- lm(...)
myrtle_lm <- lm(...)

summary(brisbane_lm)
summary(...)
```

```{r tree_model_regression-hint-3}
brisbane <- sf_trees %>% 
  filter(str_detect(species, "Brisbane Box"))

myrtle <- sf_trees %>% 
  filter(str_detect(species, "Swamp Myrtle"))


brisbane_lm <- lm(dbh ~ age, data = brisbane)
myrtle_lm <- lm(...)

summary(brisbane_lm)
summary(...)
```

```{r tree_model_regression-solution}

brisbane <- sf_trees %>% 
  filter(str_detect(species, "Brisbane Box"))

myrtle <- sf_trees %>% 
  filter(str_detect(species, "Swamp Myrtle"))


brisbane_lm <- lm(dbh ~ age, data = brisbane)
myrtle_lm <- lm(dbh ~ age, data = myrtle)

summary(brisbane_lm)
summary(myrtle_lm)
```

```{r load models into env, include=FALSE}
brisbane <- sf_trees %>% 
  filter(str_detect(species, "Brisbane Box"))

myrtle <- sf_trees %>% 
  filter(str_detect(species, "Swamp Myrtle"))


brisbane_lm <- lm(dbh ~ age, data = brisbane)
myrtle_lm <- lm(dbh ~ age, data = myrtle)

summary(brisbane_lm)
summary(myrtle_lm)
```

Discuss and interpret the models' summaries. Answer the following:

```{r grow-pace}
quiz(
  question_radio(
    "Which tree grows faster? (in terms of diameter)",
    allow_retry = T, random_answer_order = T,
    answer("On average, Brisbane Box grows twice as fast as Swamp Myrtle", correct = T),
    answer("On average, Swamp Myrtle grows twice as fast as Brisbane Box"),
    answer("The answer cannot be deduced from the models' summaries, further analysis should be done"),
    answer("Both trees grow in almost the same pace")
  )
)
```

## Validating regression assumptions

In the previous section we created two linear models. Using a proper plot examine the first model's residuals (the Brisbane Box model) with a qqplot and also with a scatter plot (residual as a function of tree age). Do linear assumptions hold? Which assumptions are violated?

To assist you with the exercise, here are the two charts which you should get by the end of this section:

```{r validating example of charts, echo = FALSE, fig.height=3, fig.width=6}
plot1 <- brisbane %>% 
  mutate(bris_resid = brisbane_lm$residuals) %>% 
  ggplot(aes(sample = bris_resid)) +
  geom_qq() + 
  ggtitle("QQ-plot of residuals")

plot2 <- brisbane %>% 
  mutate(bris_resid = brisbane_lm$residuals) %>% 
  ggplot(aes(y = bris_resid, x = age)) +
  geom_point() + 
  ggtitle("Residuals as a function of age")

cowplot::plot_grid(plot1, plot2)
```


```{r linear-regression-assumptions, exercise = TRUE}
# qqplot:
brisbane %>% 
  mutate(bris_resid = ...) %>% 
  ggplot(aes(... = ...)) +
  geom_...()

# scatter plot:
brisbane %>% 
  mutate(bris_resid = ...) %>% 
  ggplot(aes(y = ..., x = ...)) +
  geom_...()
```

```{r linear-regression-assumptions-hint-1}
# qqplot:
brisbane %>% 
  mutate(bris_resid = brisbane_lm$residuals) %>% 
  ggplot(aes(... = ...)) +
  geom_...()

# scatter plot:
brisbane %>% 
  mutate(bris_resid = brisbane_lm$residuals) %>% 
  ggplot(aes(y = ..., x = ...)) +
  geom_...()
```

```{r linear-regression-assumptions-hint-2}
# qqplot:
brisbane %>% 
  mutate(bris_resid = brisbane_lm$residuals) %>% 
  ggplot(aes(sample = bris_resid)) +
  geom_...()

# scatter plot:
brisbane %>% 
  mutate(bris_resid = brisbane_lm$residuals) %>% 
  ggplot(aes(y = bris_resid, x = ...)) +
  geom_...()
```

```{r linear-regression-assumptions-solution}
brisbane %>% 
  mutate(bris_resid = brisbane_lm$residuals) %>% 
  ggplot(aes(sample = bris_resid)) +
  geom_qq()

brisbane %>% 
  mutate(bris_resid = brisbane_lm$residuals) %>% 
  ggplot(aes(y = bris_resid, x = age)) +
  geom_point()

```



## Bonus: plotting interactive maps (with the leaflet package)

This part is not related to linear regression, however, our data has location coordinates, so this is a nice opportunity to show the `leaflet` package.

This is not something you will be required to do in the project or the test, just good knowledge to have.

First, create a tibble called `rare_tree` which includes 50 observations of the rarest tree species in SF.

```{r prep-leaflet-tibble, exercise=TRUE}

rare_sf_trees <- sf_trees %>% 
  add_count(...) %>% 
  arrange(...) %>% 
  slice(...)

```

```{r prep-leaflet-tibble-solution}

rare_sf_trees <- sf_trees %>% 
  add_count(species) %>% 
  arrange(n) %>% 
  slice(1:50)

```

```{r save prep-leaflet-tibble}

rare_sf_trees <- sf_trees %>% 
  add_count(species) %>% 
  arrange(n) %>% 
  slice(1:50)

```

Use the following code and `leaflet` vignette to render an interactive map of these trees. You should map the longitude and latitude to the proper function parameters (of `addMarkers`) and also map a label (i.e., the tree species). Beware to only use the filtered version (i.e., `rare_sf_trees` with the 50 trees, otherwise you might crash the server).

```{r leaflet-tree-map, exercise=TRUE}

library(leaflet)

rare_sf_trees %>% 
  leaflet() %>% 
  addTiles() %>% 
  addMarkers(lng = ~...,
             lat = ~...,
             label = ~...)

```



```{r leaflet-tree-map-solution}
library(leaflet)

rare_sf_trees %>% 
  leaflet() %>% 
  addTiles() %>% 
  addMarkers(lng = ~longitude,
             lat = ~latitude,
             label = ~species)

```
